Re-Examine Hybrid State-Space and Self-Attention for In-Context Learning

### 2025-01-26 paper.tex
change introduction

% 在自然语言处理（NLP）领域，SSM被用于建模文本序列的动态特性。其优势在于能够高效地处理长序列数据，并通过状态变量的更新机制捕捉序列中的时间依赖性。然而，传统的SSM存在一些局限性，例如线性时不变（LTI）系统的矩阵参数（如A、B、C）在整个序列生成过程中保持固定，这使得模型难以进行内容感知的推理。例如，在处理需要选择性关注或忽略特定输入的任务时，传统SSM的表现不如基于自注意力机制的Transformer模型。
In the field of natural language processing (NLP), SSM is used to model the dynamic characteristics of text sequences. Its advantage is that it can efficiently handle long sequence data and capture the time dependency in the sequence through the update mechanism of the state variables. However, traditional SSM has some limitations, such as the matrix parameters (such as A, B, C) of the linear time-invariant (LTI) system remain fixed throughout the entire sequence generation process, making it difficult for the model to perform content-aware reasoning. For example, when dealing with tasks that require selective attention or ignoring specific inputs, the performance of traditional SSM is not as good as the Transformer model based on the self-attention mechanism.

\noindent
\begin{minipage}{0.48\textwidth}
\small
\begin{align}
   h'(t) &= A h(t) + B x(t) \\
   y(t) &= C h(t)
\end{align}
\end{minipage}
\begin{minipage}{0.48\textwidth}
\small
\begin{align}
   h'(t) &= A h(t) + B x(t) \\
   y(t) &= C h(t)
\end{align}
\end{minipage}

\noindent
\begin{minipage}{0.48\textwidth}
\small
\begin{align}
   h'(t) &= A h(t) + B x(t) \\
   y(t) &= C h(t)
\end{align}
\end{minipage}
\begin{minipage}{0.48\textwidth}
\small
\begin{align}
   h'(t) &= A h(t) + B x(t) \\
   y(t) &= C h(t)
\end{align}
\end{minipage}

%为了克服这些局限性，研究者们提出了多种改进方法。例如，~\cite{mamba2}模型通过引入选择性扫描算法（Selective Scan）和硬件感知算法（Hardware-Aware Algorithm），使得SSM能够动态调整参数，从而更好地处理长序列数据并提高计算效率。这种改进的SSM，也被称为选择性状态空间模型（Selective SSM，简称S6），追求能够像Transformer一样构建高效的序列处理模块。但是即便如此，它仍然不如CNN 和 Transformer在计算上高效，而现代并行计算设备的发展使得这种计算效率的差距变得更加明显, 从而限制了SSM在更大规模的序列数据上的训练,因此揭示选择性SSM和注意力机制之间的更深层次关系, 并提出一种新的混合建模架构, 以克服这些局限性是非常重要的.

To overcome these limitations, researchers have proposed various improvement methods. For example, the Mamba~\cite{gu2023mamba} model introduces the Selective Scan algorithm and the Hardware-Aware Algorithm, allowing SSM to dynamically adjust parameters to better handle long sequence data and improve computational efficiency. This improved SSM, also known as Selective State Space Model (Selective SSM, abbreviated as S6), aims to build an efficient sequence processing module like Transformer. However, even so, it is still not as computationally efficient as CNN and Transformer, and the development of modern parallel computing devices makes this computational efficiency gap more pronounced, limiting the training of SSM on larger-scale sequence data. Therefore, it is very important to reveal a deeper relationship between Selective SSM and the attention mechanism and propose a new hybrid modeling architecture to overcome these limitations.

>>>>

% 在自然语言处理（NLP）领域，SSM被用于建模文本序列的动态特性。其优势在于能够高效地处理长序列数据，并通过状态变量的更新机制捕捉序列中的时间依赖性。然而，传统的SSM存在一些局限性，例如线性时不变（LTI）系统的矩阵参数（如A、B、C）在整个序列生成过程中保持固定，这使得模型难以进行内容感知的推理。例如，在处理需要选择性关注或忽略特定输入的任务时，传统SSM的表现不如基于自注意力机制的Transformer模型。
      In the field of natural language processing (NLP), SSM is used to model the dynamic characteristics of text sequences. Its advantage is that it can efficiently handle long sequence data and capture the time dependency in the sequence through the update mechanism of the state variables. However, traditional SSM has some limitations, such as the matrix parameters (such as A, B, C) of the linear time-invariant (LTI) system remain fixed throughout the entire sequence generation process, making it difficult for the model to perform content-aware reasoning. For example, when dealing with tasks that require selective attention or ignoring specific inputs, the performance of traditional SSM is not as good as the Transformer model based on the self-attention mechanism.
      % 状态空间模型是一种用于描述系统动态行为的数学模型，其基本形式为：
      State-space models are mathematical models used to describe the dynamic behavior of systems, and their basic form is:
      \begin{minipage}{0.45\textwidth}
         \begin{alignat}{2}
            h'(t) &= Ah(t) + Bx(t) && \tag{1a} \\
            y(t) &= Ch(t) + Dx(t) && \tag{1b}
         \end{alignat}
         %其中，h(t) 是状态向量，x(t) 是输入向量，y(t) 是输出向量，A、B、C、D 是系统矩阵 
         %翻译
         where $h(t)$ is the state vector, $x(t)$ is the input vector, $y(t)$ is the output vector, and $A$, $B$, $C$, $D$ are system matrices.

         \end{minipage}
         \hfill
         \begin{minipage}{0.45\textwidth}
         \begin{alignat}{2}
            h_t &= \overline{A}h_{t-1} + \overline{B}x_t && \tag{2a} \\
            y_t &= Ch_t && \tag{2b}
         % 从连续时间系统的状态空间模型出发，可以通过离散化方法得到离散时间的状态更新公式,方程 (2a) 和 (2b) 代表状态空间模型的离散化形式。它们描述了在离散时间间隔内，状态向量和输出向量如何基于前一个状态和当前输入进行更新。
        \end{alignat}
         Starting from the state-space model of continuous-time systems, the discrete-time state update formula can be obtained through discretization methods. Equations (2a) and (2b) represent the discrete form of the state-space model. They describe how the state vector and output vector are updated based on the previous state and current input at discrete time intervals.

         \end{minipage}
         \begin{minipage}{0.45\textwidth}
         \begin{alignat}{2}
            \overline{K} &= (C\overline{B}, CAB, \ldots, C\overline{A}^k\overline{B}, \ldots) && \tag{3a} \\
            y &= x * \overline{K} && \tag{3b}
         \end{alignat}
         \end{minipage}
         % 以上为

%以上为SSM-S4的过程，所有结构化 SSM 都是 LTI 模型，LTI 模型存在基本的效率限制,为了克服这些局限性，研究者们提出了多种改进方法。例如，~\cite{mamba2}模型通过引入选择性扫描算法（Selective Scan）和硬件感知算法（Hardware-Aware Algorithm），使得SSM能够动态调整参数，从而更好地处理长序列数据并提高计算效率。这种改进的SSM，也被称为选择性状态空间模型（Selective SSM，简称S6），追求能够像Transformer一样构建高效的序列处理模块。但是即便如此，它仍然不如CNN 和 Transformer在计算上高效，而现代并行计算设备的发展使得这种计算效率的差距变得更加明显, 从而限制了SSM在更大规模的序列数据上的训练,因此揭示选择性SSM和注意力机制之间的更深层次关系, 并提出一种新的混合建模架构, 以克服这些局限性是非常重要的.

 The above is the process of SSM-S4. All structured SSMs are LTI models, and LTI models have basic efficiency limitations. To overcome these limitations, researchers have proposed various improvement methods. For example, the Mamba~\cite{mamba2} model introduces the Selective Scan algorithm and the Hardware-Aware Algorithm, allowing SSM to dynamically adjust parameters to better handle long sequence data and improve computational efficiency. This improved SSM, also known as Selective State Space Model (Selective SSM, referred to as S6), aims to build efficient sequence processing modules like Transformer. However, even so, it is still not as efficient as CNN and Transformer in computation, and the development of modern parallel computing devices makes this computational efficiency gap more obvious, thereby limiting the training of SSM on larger-scale sequence data. Therefore, it is very important to reveal the deeper relationship between Selective SSM and the attention mechanism and propose a new hybrid modeling architecture to overcome these limitations.
